Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-12-01 response_to_reviewers.txt

45-56:

1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
* This referred to both of the cases mentioned in text (early and later
	access times). We revised the phrase to "these plots" to avoid
	confusion.

2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
* These plots demonstrate the effect of curvature on access times when variables such as degree are controlled. Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.
	A pair of trees that both contain leaves 1 and 2 as siblings are an
	example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves.

**** EM We need to revise the text including your response rather than just respond.

@@

Page 17

12-15: Adding hyphens (e.g. 'distance-1 pairs') might increase the readability.
* We have made the suggested change

15-23: The traceability / comprehensibility of the statements would strongly benefit from improving the readability of the plots in Figure 5 and 6.
* TODO
*** EM Perhaps we can respond by just saying that all of the statements regarding the influence of curvature are confirmed by the regression analysis.

@@

53: It looks like there is too much white space behind 'w.l.o.g.'.
* Corrected.

Page 18

Proof of (i):The proof only works when the '+14' in the numerator is not reduced to '+12' (which is also not necessary, because bounding the sum of the logarithms in the denominator is sufficient.
* TODO: Do you see a problem here Erick? Reducing the numerator from 14
	to 12 is valid. Maybe he is confused by the last step, where I borrow
	from the reduced -n in the denominator to reduce the 20 to 18?
**** EM: I'm a little confused by the very last step. Don't we want the 18 to be a 16?

@@

2. In the phrase 'where one of the root's children is small', is the intended meaning exactly one or at least one? Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the majority of) the leaves), the statement does not hold in general (neither for exactly nor for at least). Consider, e.g., a tree with a leaf as the only small child of the root and at least three small subtrees with two leaves (for the case exactly) respectively the balanced tree with four leaves (for the case at least).
* See the previous response.

36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more consistent with the previous use of numerals).
* We have made the suggested change.

38-42:

1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n^2-4n-6) is not correct for all intended n. The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves. Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
* Thank you for pointing this out. To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.

2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case distinction n = 3 vs. n ≥ 4).
* As described above we restrict to n>3.

@@

45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
* We added the statement to line 45.

54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3 for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).  This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
* There appears to be some confusion on what constitutes a moved subtree here.  From our distribution of rSPR moves, the balanced tree with n = 4 leaves does not move any subtrees of size 2. There are moves of such trees that are equivalent to moves of leaves which appears to be the source of the confusion. As such, the average moved subtree size for n = 8 is 1.4 for the ladder tree and approximately 1.379 (160/116) for the balanced tree. We have added the following explanation to make this more clear to the reader:

"Now, we bound the average size of the moved subtrees (in terms of the number of leaves) from rSPR moves according to our distribution. Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees. For example, from the balanced tree with 4 leaves, we do not consider moves of the root’s 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution."

### 2016-12-01 response_to_reviewers.txt

****TODO FIGURES END*****

45-56:

1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
* This referred to both of the cases mentioned in text (early and later
	access times). We revised the phrase to "these plots" to avoid
	confusion.

2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
* These plots demonstrate the effect of curvature on access times when variables such as degree are controlled. Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.  A pair of trees that both contain leaves 1 and 2 as siblings are an example of inclusions of the pair of trees obtained by merging those siblings into a single leaf and relabeling the leaves. We rephrased this sentence to state "To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g. pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."

@@

Page 21

49:

1. The meaning of the function L(.) can be guessed from the context without too much effort, but providing a short definition / explanation would help avoiding any misconceptions.
* We added "... , where L(Q) is the label set of Q, ..." to the sentence.

2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
* Yes, because of the triangle formed by T, S, and U, as we state in the proof. T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position. We now state: In the latter case, S is reached from U by moving Q in a manner that displaces T from its original position, because T and S differ only in the location of R and d_SPR(T, U ) = d_SPR(S, U ) = 1. Thus, T |(X \ L(Q)) = S|(X \ L(Q)), where L(Q) is the label set of Q."

### 2016-12-01 curvature.tex
fixed Lemma 5.2 proof and response
fixed line breaks in response document. Added a couple review comments to the text

\begin{lem}
	\label{lem:degree_max_delta}
	Let $T$,$S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lem}
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3, \\
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		&=\ 4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

>>>>

\begin{lem}
	\label{lem:degree_max_delta}
	Let $T$,$S$ be trees with $n \ge 3$ leaves, and assume w.l.o.g. that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lem}
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 14}{4n^2 - 16n + 16 - 2(n-2)} \\
		= \ &\frac{3n^2 - 13n + 14}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 14}{4n^2 - 17\frac{1}{3}n + 18\frac{2}{3}} &\forall n \ge 3, \\
	\end{align*}
	which is at least 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		{\Delta}N &= \degree{S} - \degree{T} \\
		&\le\ (4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}) \\
		&\ \ \ \ \ \ \ \ - (3n^2 - 13n + 14) \\
		&\le\ (4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		&=\ 4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		&=\ n^2 - 5n + 6.
	\end{align*}
\end{proof}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$. 
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}
